### 2017-11-29 submission/manuscript.tex
refine major questions and problems this manuscript answers in intro

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness (P-value
\textless{} 0.05) after correcting for study effect and variable region
sequenced. Only evenness for adenoma (P-value \textless{} 0.05) resulted
in a slightly increased relative risk while lower Shannon diversity and
evenness in fecal samples resulted in a significant increase in relative
risk for carcinoma (P-value \textless{} 0.05). Previously associated
colorectal cancer genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}) followed a similar
pattern with a significantly increased relative risk by their presence
for carcinoma (P-value \textless{} 0.05) but not adenoma (P-value
\textgreater{} 0.05) with the exception of \emph{Porphyromonas} (P-value
\textless{} 0.05). Using the whole community versus only CRC associated
genera to build a prediction model resulted in higher classification
success based on Area Under the Curve (AUC) for both adenoma and
carcinoma using fecal and tissue samples. For the included studies, most
were adequately powered for large effect size differences. This may be
more amenable for carcinoma than for adenoma microbiota research due to
the smaller community level changes observed in our results.

@@

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totalling over 2229 total samples across 14 studies
{[}12--25{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope of
the previous meta-analysis to investigate whether the bacterial
community is an important risk factor for both adenoma and carcinoma. To
accomplish this we first assessed whether the diversity changes
throughout disease (control to adenoma to carcinoma) and if it results
in an increased relative risk (RR) for adenoma or carcinoma. Next, we
assessed how common CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
affect the relative risk of adenoma or carcinoma. Third, using Random
Forest models, we analyzed whether the full community or only the CRC
associated genera resulted in better model classification based on the
area under the curve (AUC). Since the changes in community were subtle
for adenoma we also examined what effect and sample size the studies
that were used were adequately powered for.

Our analysis found a continuous decrease in Shannon diversity as disease
became more severe which correlated with a significantly increased RR
for carcinoma. Using only CRC-associated genera, the RR for carcinoma
was higher relative to what was observed for the Shannon diversity RR.
Conversely, we demonstrate that the AUC of the classification models
increased when the full community was incorporated as opposed to only
CRC-associated genera. Although we analyzed data sets which sampled
large numbers of individuals, our results indicate the individual
studies were underpowered for detecting effect size differences of 10\%
or below between the case and control groups.

@@

\textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
were identified through the review articles written by Keku, \emph{et
al.} and Vogtmann, \emph{et al.} {[}26,27{]}. Additional studies not
mentioned in the reviews were obtained based on the authors' knowledge
of the literature. Studies that used tissue or feces as their sample
source for 16S rRNA gene sequencing analysis were included. Studies
using either 454 or Illumina sequencing technology were included. Only
data sets that had sequences available for analysis were included. Some
studies did not have publically available sequences or did not have
metadata in which the authors were able to share. After these filtering
steps, the following studies remained: Ahn, \emph{et al.} {[}21{]},
Baxter, \emph{et al.} {[}24{]}, Brim, \emph{et al.} {[}17{]}, Burns,
\emph{et al.} {[}22{]}, Chen, \emph{et al.} {[}14{]}, Dejea, \emph{et
al.} {[}19{]}, Flemer, \emph{et al.} {[}13{]}, Geng, \emph{et al.}
{[}25{]}, Hale, \emph{et al.} {[}12{]}, Kostic, \emph{et al.} {[}28{]},
Lu, \emph{et al.} {[}16{]}, Sanapareddy, \emph{et al.} {[}20{]}, Wang,
\emph{et al.} {[}15{]}, Weir, \emph{et al.} {[}18{]}, and Zeller,
\emph{et al.} {[}23{]}. The Zackular {[}29{]} study was not included
becasue the 90 individuals analyzed within the study are contained
within the larger Baxter study {[}24{]}. Additionally, after sequence
processing all the case samples for the Kostic study only had 100
sequences remaining and was not used. This left a total of 14 studies
for which analysis could be completed.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained from the by searching the respective accession
number of the study following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
for one study {[}21{]} and for four studies was obtained directly from
the authors {[}12,13,18,20{]}. Each study was processed using the mothur
(v1.39.3) software program {[}30{]}. Where possible, quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults, the
stated quality cut-offs were used instead. Chimeras were identifed and
removed using VSEARCH {[}31{]} before \emph{de novo} OTU clustering at
97\% similarity using the OptiClust algorithm {[}32{]} was utilized.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.2) software package {[}33{]}.
For the \(\alpha\)-diversity analysis, values were power transformed
using the rcompanion (v1.10.1) package {[}34{]} and then Z-score
normalized using the car (v2.1.5) package {[}35{]}. Testing for
\(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.14) package {[}36{]} to correct for study
and variable region effects in feces and study, variable region, and
individual effects in tissue. Relative risk was analyzed using both the
epiR (v0.9.87) and metafor (v2.0.0) packages {[}37,38{]} by assessing
how many with and without disease were above and below the overall
median value within the specific study. Relative risk significance
testing utilized the chi-squred test. \(\beta\)-diversity differences
utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
vegan (v2.4.4) package {[}39{]}. Random Forest models were built using
both the caret (v6.0.77) and randomForest (v4.6.12) packages
{[}40,41{]}. Differences between the obtained AUC versus a random model
AUC was assessed using T-tests. Power analysis and estimations were made
using the pwr (v1.2.1) and statmod (v1.4.30) packages {[}42,43{]}. All
figures were created using both ggplot2 (v2.2.1) and gridExtra (v2.3)
packages {[}44,45{]}.

@@

\hypertarget{ref-hale_shifts_2017}{}
12. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
Cancer Epidemiology, Biomarkers \& Prevention: A Publication of the
American Association for Cancer Research, Cosponsored by the American
Society of Preventive Oncology. 2017;26:85--94.

\hypertarget{ref-flemer_tumour-associated_2017}{}
13. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
al. Tumour-associated and non-tumour-associated microbiota in colorectal
cancer. Gut. 2017;66:633--43.

\hypertarget{ref-chen_human_2012}{}
14. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
mucosa-associated microbiota in patients with colorectal cancer. PloS
One. 2012;7:e39743.

\hypertarget{ref-wang_structural_2012}{}
15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
segregation of gut microbiota between colorectal cancer patients and
healthy volunteers. The ISME journal. 2012;6:320--9.

\hypertarget{ref-lu_mucosal_2016}{}
16. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
adherent bacterial dysbiosis in patients with colorectal adenomas.
Scientific Reports. 2016;6:26337.

\hypertarget{ref-brim_microbiome_2013}{}
17. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
al. Microbiome analysis of stool samples from African Americans with
colon polyps. PloS One. 2013;8:e81352.

\hypertarget{ref-weir_stool_2013}{}
18. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
Stool microbiome and metabolome differences between colorectal cancer
patients and healthy adults. PloS One. 2013;8:e70803.

\hypertarget{ref-dejea_microbiota_2014}{}
19. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, et al. Microbiota organization is a distinct feature of
proximal colorectal cancers. Proceedings of the National Academy of
Sciences of the United States of America. 2014;111:18321--6.

\hypertarget{ref-sanapareddy_increased_2012}{}
20. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
et al. Increased rectal microbial richness is associated with the
presence of colorectal adenomas in humans. The ISME journal.
2012;6:1858--68.

\hypertarget{ref-ahn_human_2013}{}
21. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
microbiome and risk for colorectal cancer. Journal of the National
Cancer Institute. 2013;105:1907--11.

\hypertarget{ref-burns_virulence_2015}{}
22. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
are a signature of the microbiome in the colorectal tumor
microenvironment. Genome Medicine. 2015;7:55.

\hypertarget{ref-zeller_potential_2014}{}
23. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
Potential of fecal microbiota for early-stage detection of colorectal
cancer. Molecular Systems Biology. 2014;10:766.

\hypertarget{ref-baxter_microbiota-based_2016}{}
24. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
improves the sensitivity of fecal immunochemical test for detecting
colonic lesions. Genome Medicine. 2016;8:37.

\hypertarget{ref-geng_diversified_2013}{}
25. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.

\hypertarget{ref-keku_gastrointestinal_2015}{}
26. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
microbiota and colorectal cancer. American Journal of Physiology -
Gastrointestinal and Liver Physiology {[}Internet{]}. 2015 {[}cited 2017
Oct 30{]};308:G351--63. Available from:
\url{http://ajpgi.physiology.org/lookup/doi/10.1152/ajpgi.00360.2012}

\hypertarget{ref-vogtmann_epidemiologic_2016}{}
27. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
microbiome and cancer. British Journal of Cancer {[}Internet{]}. 2016
{[}cited 2017 Oct 30{]};114:237--42. Available from:
\url{http://www.nature.com/doifinder/10.1038/bjc.2015.465}

\hypertarget{ref-kostic_genomic_2012}{}
28. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
al. Genomic analysis identifies association of Fusobacterium with
colorectal carcinoma. Genome Research. 2012;22:292--8.

\hypertarget{ref-zackular_human_2014}{}
29. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
microbiome as a screening tool for colorectal cancer. Cancer Prevention
Research (Philadelphia, Pa.). 2014;7:1112--21.

\hypertarget{ref-schloss_introducing_2009}{}
30. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
EB, et al. Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. Appl.Environ.Microbiol. {[}Internet{]}. 2009 {[}cited 12AD
Jan 1{]};75:7537--41. Available from:
\url{http://aem.asm.org/cgi/content/abstract/75/23/7537}

\hypertarget{ref-rognes_vsearch_2016}{}
31. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.

\hypertarget{ref-westcott_opticlust_2017}{}
32. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
Amplicon-Based Sequence Data to Operational Taxonomic Units. mSphere.
2017;2.

\hypertarget{ref-r_citation_2017}{}
33. R Core Team. R: A language and environment for statistical computing
{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
2017. Available from: \url{https://www.R-project.org/}

\hypertarget{ref-rcompanion_citation_2017}{}
34. Mangiafico S. Rcompanion: Functions to support extension education
program evaluation {[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=rcompanion}

\hypertarget{ref-car_citation_2011}{}
35. Fox J, Weisberg S. An R companion to applied regression
{[}Internet{]}. Second. Thousand Oaks CA: Sage; 2011. Available from:
\url{http://socserv.socsci.mcmaster.ca/jfox/Books/Companion}

\hypertarget{ref-lme4_citation_2015}{}
36. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
models using lme4. Journal of Statistical Software. 2015;67:1--48.

\hypertarget{ref-epir_citation_2017}{}
37. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
J, Thornton R, Reiczigel J, et al. EpiR: Tools for the analysis of
epidemiological data {[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=epiR}

\hypertarget{ref-metafor_citation_2010}{}
38. Viechtbauer W. Conducting meta-analyses in R with the metafor
package. Journal of Statistical Software {[}Internet{]}. 2010;36:1--48.
Available from: \url{http://www.jstatsoft.org/v36/i03/}

\hypertarget{ref-vegan_citation_2017}{}
39. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
et al. Vegan: Community ecology package {[}Internet{]}. 2017. Available
from: \url{https://CRAN.R-project.org/package=vegan}

\hypertarget{ref-caret_citation_2017}{}
40. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
Cooper T, et al. Caret: Classification and regression training
{[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=caret}

\hypertarget{ref-randomforest_citation_2002}{}
41. Liaw A, Wiener M. Classification and regression by randomForest. R
News {[}Internet{]}. 2002;2:18--22. Available from:
\url{http://CRAN.R-project.org/doc/Rnews/}

\hypertarget{ref-pwr_citation_2017}{}
42. Champely S. Pwr: Basic functions for power analysis {[}Internet{]}.
2017. Available from: \url{https://CRAN.R-project.org/package=pwr}

\hypertarget{ref-statmod_citation_2016}{}
43. Giner G, Smyth GK. Statmod: Probability calculations for the inverse
gaussian distribution. R Journal. 2016;8:339--51.

\hypertarget{ref-ggplot2_citation_2009}{}
44. Wickham H. Ggplot2: Elegant graphics for data analysis
{[}Internet{]}. Springer-Verlag New York; 2009. Available from:
\url{http://ggplot2.org}

\hypertarget{ref-gridextra_citation_2017}{}
45. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
{[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=gridExtra}

>>>>

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness (P-value
\textless{} 0.05) after correcting for study effect and variable region
sequenced. Only evenness for adenoma (P-value \textless{} 0.05) resulted
in a slightly increased relative risk while lower Shannon diversity and
evenness in fecal samples resulted in a significant increase in relative
risk for carcinoma (P-value \textless{} 0.05). Previously associated
colorectal cancer genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}) followed a similar
pattern with a significantly increased relative risk by their presence
for carcinoma (P-value \textless{} 0.05) but not adenoma (P-value
\textgreater{} 0.05) with the exception of \emph{Porphyromonas} (P-value
\textless{} 0.05). Using the whole community versus only CRC associated
genera to build a prediction model resulted in higher classification
success based on Area Under the Curve (AUC) for both adenoma and
carcinoma using fecal and tissue samples. The most important OTUs for
these models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. For the included studies, most were adequately powered for
large effect size differences. This may be more amenable for carcinoma
than for adenoma microbiota research due to the smaller community level
changes observed in our results.

@@

The focus on biomarkers within stool makes sense since the current gold
standard for diagnosis is a colonoscopy which can be time-consuming and
is not without risks. Stool offers an easy and cost-effective way in
which to stratify risk within the population. Although stool represents
an easy and less invasive way to assess risk, it is not clear how
reflective this sample is to what is actually happening on the adenoma
or carcinoma. Some studies have begun to try and assess this in health
and disease but are limited by their sample size {[}12,13{]}. Since the
stool possibly represents an agregate community of the overall
gastrointestinal (GI) tract, sampling the adenoma or carinoma directly
might do a better job at identifying important microbes and could be
drastically different then what is identified in stool. Sampling the
tissue directly may provide clearer answers but it is not without
problems. Due to bowel prep the communities left for sampling may not be
the truly resident microbes but rather a collection of what is able to
keep adhered to the mucosa. Further, these samples would contain much
more host DNA then microbial DNA. Low biomass samples are very difficult
to work with and results can end up being study dependent due to the
randomness of contamination {[}14{]}. Due to these variables, one
question our meta-analysis aims to answer is whether it is possible to
find consistent patterns that emerge across studies regardless of
whether they used stool or tissue samples.

This intense focus on identifying biomarkers has identified mostly
mouth-associated microbes as potential CRC-associated microbes in the GI
tract {[}15--17{]}. The main bacteria of interest arising from this set
of microbes has been those within the \emph{Fusobacterium} genus. Yet,
the question remains as to whether or not this is indeed the most
important genera to be focusing on since many studies microbiota based
studies identify a collection of oral microbes rather than single genera
{[}16,17{]}. Based on this discrepency, the second question we want to
answer with this meta-analysis is if there really is one dominant
CRC-associated genera that can be identified across studies.

The heavy focus on the identification of biomarkers has mostly dominated
the study of CRC and has had an unattended consequence of reducing the
focus on changes that occur within the underlying resident community.
This has been borne out by the majority of previous studies, both within
stool and tissue, that have focused predominately on biomarkers as well
as the only meta-analysis investigating this area {[}11{]}. Since it has
not been investigated in detail the final question that this
meta-analysis aims to answer is whether there are consistent detectable
community differences as disease severity increases.

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totalling over 2229 total samples across 14 studies
{[}12,16--28{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope
of the previous meta-analysis to investigate whether the bacterial
community is an important risk factor for both adenoma and carcinoma. To
accomplish this we first assessed whether the diversity changes
throughout disease (control to adenoma to carcinoma) and if it results
in an increased relative risk (RR) for adenoma or carcinoma. Next, we
assessed how common CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
affect the relative risk of adenoma or carcinoma. Third, using Random
Forest models, we analyzed whether the full community or only the CRC
associated genera resulted in better model classification based on the
area under the curve (AUC). We observed that the community changes as
disease severity worsens and that this community is important for
disease classification. However, since the changes in community were
subtle for adenoma we also examined what effect and sample size the
studies that were used were adequately powered for. Although we analyzed
data sets which sampled large numbers of individuals, our results
indicate the individual studies were underpowered for detecting effect
size differences of 10\% or below between the case and control groups.

@@

\textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
were identified through the review articles written by Keku, \emph{et
al.} and Vogtmann, \emph{et al.} {[}29,30{]}. Additional studies not
mentioned in the reviews were obtained based on the authors' knowledge
of the literature. Studies that used tissue or feces as their sample
source for 16S rRNA gene sequencing analysis were included. Studies
using either 454 or Illumina sequencing technology were included. Only
data sets that had sequences available for analysis were included. Some
studies did not have publically available sequences or did not have
metadata in which the authors were able to share. After these filtering
steps, the following studies remained: Ahn, \emph{et al.} {[}26{]},
Baxter, \emph{et al.} {[}16{]}, Brim, \emph{et al.} {[}22{]}, Burns,
\emph{et al.} {[}27{]}, Chen, \emph{et al.} {[}19{]}, Dejea, \emph{et
al.} {[}24{]}, Flemer, \emph{et al.} {[}12{]}, Geng, \emph{et al.}
{[}28{]}, Hale, \emph{et al.} {[}18{]}, Kostic, \emph{et al.} {[}15{]},
Lu, \emph{et al.} {[}21{]}, Sanapareddy, \emph{et al.} {[}25{]}, Wang,
\emph{et al.} {[}20{]}, Weir, \emph{et al.} {[}23{]}, and Zeller,
\emph{et al.} {[}17{]}. The Zackular {[}31{]} study was not included
becasue the 90 individuals analyzed within the study are contained
within the larger Baxter study {[}16{]}. Additionally, after sequence
processing all the case samples for the Kostic study only had 100
sequences remaining and was not used. This left a total of 14 studies
for which analysis could be completed.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained from the by searching the respective accession
number of the study following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
for one study {[}26{]} and for four studies was obtained directly from
the authors {[}12,18,23,25{]}. Each study was processed using the mothur
(v1.39.3) software program {[}32{]}. Where possible, quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults, the
stated quality cut-offs were used instead. Chimeras were identifed and
removed using VSEARCH {[}33{]} before \emph{de novo} OTU clustering at
97\% similarity using the OptiClust algorithm {[}34{]} was utilized.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.2) software package {[}35{]}.
For the \(\alpha\)-diversity analysis, values were power transformed
using the rcompanion (v1.10.1) package {[}36{]} and then Z-score
normalized using the car (v2.1.5) package {[}37{]}. Testing for
\(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.14) package {[}38{]} to correct for study
and variable region effects in feces and study, variable region, and
individual effects in tissue. Relative risk was analyzed using both the
epiR (v0.9.87) and metafor (v2.0.0) packages {[}39,40{]} by assessing
how many with and without disease were above and below the overall
median value within the specific study. Relative risk significance
testing utilized the chi-squred test. \(\beta\)-diversity differences
utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
vegan (v2.4.4) package {[}41{]}. Random Forest models were built using
both the caret (v6.0.77) and randomForest (v4.6.12) packages
{[}42,43{]}. Differences between the obtained AUC versus a random model
AUC was assessed using T-tests. Power analysis and estimations were made
using the pwr (v1.2.1) and statmod (v1.4.30) packages {[}44,45{]}. All
figures were created using both ggplot2 (v2.2.1) and gridExtra (v2.3)
packages {[}46,47{]}.

@@

\hypertarget{ref-flemer_tumour-associated_2017}{}
12. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
al. Tumour-associated and non-tumour-associated microbiota in colorectal
cancer. Gut. 2017;66:633--43.

\hypertarget{ref-Flynn_preprint_2017}{}
13. Flynn KJ, Ruffin MT, Turgeon DK, Schloss PD. Spatial variation of
the native colon microbiota in healthy adults. Cold Spring Harbor
Laboratory; 2017; Available from: \url{https://doi.org/10.1101/189886}

\hypertarget{ref-Salter_contamination_2014}{}
14. Salter SJ, Cox MJ, Turek EM, Calus ST, Cookson WO, Moffatt MF, et
al. Reagent and laboratory contamination can critically impact
sequence-based microbiome analyses. BMC Biology {[}Internet{]}. Springer
Nature; 2014;12. Available from:
\url{https://doi.org/10.1186/s12915-014-0087-z}

\hypertarget{ref-kostic_genomic_2012}{}
15. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
al. Genomic analysis identifies association of Fusobacterium with
colorectal carcinoma. Genome Research. 2012;22:292--8.

\hypertarget{ref-baxter_microbiota-based_2016}{}
16. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
improves the sensitivity of fecal immunochemical test for detecting
colonic lesions. Genome Medicine. 2016;8:37.

\hypertarget{ref-zeller_potential_2014}{}
17. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
Potential of fecal microbiota for early-stage detection of colorectal
cancer. Molecular Systems Biology. 2014;10:766.

\hypertarget{ref-hale_shifts_2017}{}
18. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
Cancer Epidemiology, Biomarkers \& Prevention: A Publication of the
American Association for Cancer Research, Cosponsored by the American
Society of Preventive Oncology. 2017;26:85--94.

\hypertarget{ref-chen_human_2012}{}
19. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
mucosa-associated microbiota in patients with colorectal cancer. PloS
One. 2012;7:e39743.

\hypertarget{ref-wang_structural_2012}{}
20. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
segregation of gut microbiota between colorectal cancer patients and
healthy volunteers. The ISME journal. 2012;6:320--9.

\hypertarget{ref-lu_mucosal_2016}{}
21. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
adherent bacterial dysbiosis in patients with colorectal adenomas.
Scientific Reports. 2016;6:26337.

\hypertarget{ref-brim_microbiome_2013}{}
22. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
al. Microbiome analysis of stool samples from African Americans with
colon polyps. PloS One. 2013;8:e81352.

\hypertarget{ref-weir_stool_2013}{}
23. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
Stool microbiome and metabolome differences between colorectal cancer
patients and healthy adults. PloS One. 2013;8:e70803.

\hypertarget{ref-dejea_microbiota_2014}{}
24. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, et al. Microbiota organization is a distinct feature of
proximal colorectal cancers. Proceedings of the National Academy of
Sciences of the United States of America. 2014;111:18321--6.

\hypertarget{ref-sanapareddy_increased_2012}{}
25. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
et al. Increased rectal microbial richness is associated with the
presence of colorectal adenomas in humans. The ISME journal.
2012;6:1858--68.

\hypertarget{ref-ahn_human_2013}{}
26. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
microbiome and risk for colorectal cancer. Journal of the National
Cancer Institute. 2013;105:1907--11.

\hypertarget{ref-burns_virulence_2015}{}
27. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
are a signature of the microbiome in the colorectal tumor
microenvironment. Genome Medicine. 2015;7:55.

\hypertarget{ref-geng_diversified_2013}{}
28. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.

\hypertarget{ref-keku_gastrointestinal_2015}{}
29. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
microbiota and colorectal cancer. American Journal of Physiology -
Gastrointestinal and Liver Physiology {[}Internet{]}. 2015 {[}cited 2017
Oct 30{]};308:G351--63. Available from:
\url{http://ajpgi.physiology.org/lookup/doi/10.1152/ajpgi.00360.2012}

\hypertarget{ref-vogtmann_epidemiologic_2016}{}
30. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
microbiome and cancer. British Journal of Cancer {[}Internet{]}. 2016
{[}cited 2017 Oct 30{]};114:237--42. Available from:
\url{http://www.nature.com/doifinder/10.1038/bjc.2015.465}

\hypertarget{ref-zackular_human_2014}{}
31. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
microbiome as a screening tool for colorectal cancer. Cancer Prevention
Research (Philadelphia, Pa.). 2014;7:1112--21.

\hypertarget{ref-schloss_introducing_2009}{}
32. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
EB, et al. Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. Appl.Environ.Microbiol. {[}Internet{]}. 2009 {[}cited 12AD
Jan 1{]};75:7537--41. Available from:
\url{http://aem.asm.org/cgi/content/abstract/75/23/7537}

\hypertarget{ref-rognes_vsearch_2016}{}
33. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.

\hypertarget{ref-westcott_opticlust_2017}{}
34. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
Amplicon-Based Sequence Data to Operational Taxonomic Units. mSphere.
2017;2.

\hypertarget{ref-r_citation_2017}{}
35. R Core Team. R: A language and environment for statistical computing
{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
2017. Available from: \url{https://www.R-project.org/}

\hypertarget{ref-rcompanion_citation_2017}{}
36. Mangiafico S. Rcompanion: Functions to support extension education
program evaluation {[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=rcompanion}

\hypertarget{ref-car_citation_2011}{}
37. Fox J, Weisberg S. An R companion to applied regression
{[}Internet{]}. Second. Thousand Oaks CA: Sage; 2011. Available from:
\url{http://socserv.socsci.mcmaster.ca/jfox/Books/Companion}

\hypertarget{ref-lme4_citation_2015}{}
38. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
models using lme4. Journal of Statistical Software. 2015;67:1--48.

\hypertarget{ref-epir_citation_2017}{}
39. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
J, Thornton R, Reiczigel J, et al. EpiR: Tools for the analysis of
epidemiological data {[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=epiR}

\hypertarget{ref-metafor_citation_2010}{}
40. Viechtbauer W. Conducting meta-analyses in R with the metafor
package. Journal of Statistical Software {[}Internet{]}. 2010;36:1--48.
Available from: \url{http://www.jstatsoft.org/v36/i03/}

\hypertarget{ref-vegan_citation_2017}{}
41. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
et al. Vegan: Community ecology package {[}Internet{]}. 2017. Available
from: \url{https://CRAN.R-project.org/package=vegan}

\hypertarget{ref-caret_citation_2017}{}
42. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
Cooper T, et al. Caret: Classification and regression training
{[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=caret}

\hypertarget{ref-randomforest_citation_2002}{}
43. Liaw A, Wiener M. Classification and regression by randomForest. R
News {[}Internet{]}. 2002;2:18--22. Available from:
\url{http://CRAN.R-project.org/doc/Rnews/}

\hypertarget{ref-pwr_citation_2017}{}
44. Champely S. Pwr: Basic functions for power analysis {[}Internet{]}.
2017. Available from: \url{https://CRAN.R-project.org/package=pwr}

\hypertarget{ref-statmod_citation_2016}{}
45. Giner G, Smyth GK. Statmod: Probability calculations for the inverse
gaussian distribution. R Journal. 2016;8:339--51.

\hypertarget{ref-ggplot2_citation_2009}{}
46. Wickham H. Ggplot2: Elegant graphics for data analysis
{[}Internet{]}. Springer-Verlag New York; 2009. Available from:
\url{http://ggplot2.org}

\hypertarget{ref-gridextra_citation_2017}{}
47. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
{[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=gridExtra}
